---
title: "Reading 📖" 
toc: false
---

---

If you want to know why I created this list, check [I'm starting a reading list!](posts/2024-02-01_starting_a_reading_list/index.qmd).

If you want my opinion on some title, make any recommendation, or simply discuss about some book, just reach me out by e-mail.

---

```{r}
#| echo: false
library(htmltools)

generate_book_card <- function(title, subtitle, author, img_source) {
    tags$div(
        class = "book-card",
        tags$img(
            src = img_source
            #, style = "border: solid 1px red;"
        ),
        tags$div(
            class = "book-card-content",
            #style = "border: solid 1px red ;",
            tags$div(
                tags$h3(title, class = "card-title"),
                tags$p(subtitle),
                class = "book-header"
            ),
            tags$p(author, class = "book-author")
        )
    )
}
data <- jsonlite::read_json(here::here("data", "readings.json"))
```


## In progress


```{r}
#| echo: false
#| results: asis
data_wip <- list()
for (i in seq_along(data)) {
    entry <- data[[i]]
    if (entry$year_read == "wip") {
        data_wip <- c(data_wip, list(entry))
    }
}

output <- lapply(
    data_wip,
    function(entry) {
        as.character(
            generate_book_card(
                entry$title,
                entry$subtitle,
                entry$author,
                paste0("data/books/", entry$cover)
            )
        )
    }
)

output <- c("<div class = 'books-listing'>", output, "</div>")
cat(
    "```{=html}\n", 
    paste(output, collapse="\n"), 
    "\n```"
)
```


## General

```{r}
#| echo: false
#| results: asis

data_done <- list()
for (i in seq_along(data)) {
    entry <- data[[i]]
    if (entry$year_read != "wip") {
        data_done <- c(data_done, list(entry))
    }
}

output <- lapply(
    data_done,
    function(entry) {
        as.character(
            generate_book_card(
                entry$title,
                entry$subtitle,
                entry$author,
                paste0("data/books/", entry$cover)
            )
        )
    }
)

output <- c("<div class = 'books-listing'>", output, "</div>")
cat(
    "```{=html}\n", 
    paste(output, collapse="\n"), 
    "\n```"
)
```

## Technical

I rarely read statistics and technical books from cover to cover, so I'm not going to 
pretend that I've completely covered all of the books listed here.
Typically, I focus on specific chapters that are of my interest, although there are instances
where I read the entire book. Additionally, I often find myself revisiting some of these
books for a second read.

- [Functional programming in R](https://link.springer.com/book/10.1007/978-1-4842-2746-6).<br> Thomas Mailund (2017)
- [Engineering production-grade Shiny apps](https://www.routledge.com/Engineering-Production-Grade-Shiny-Apps/Fay-Rochette-Guyader-Girard/p/book/9780367466022). <br> Colin Fay, Sébastien Rochette, Vincent Guyader and Cervan Girard (2022)
- [Advanced R](https://www.routledge.com/Advanced-R-Second-Edition/Wickham/p/book/9780815384571). <br> Hadley Wickham (2019)
- [Extending R](https://www.routledge.com/Extending-R/Chambers/p/book/9781498775717). <br> John M. Chambers (2016)
- [R Graphics](https://www.routledge.com/R-Graphics-Third-Edition/Murrell/p/book/9780367780692). <br> Paul Murrell (2019)
- [Crafting Interpreters](https://craftinginterpreters.com/). <br> Robert Nystrom (2021)
- [Bayesian Data Analysis](http://www.stat.columbia.edu/~gelman/book/). <br> Andrew Gelman, John Carlin, Hal Stern, David Dunson, Aki Vehtari, and Donald Rubin (2013)
- [The Elements of Statistical Learning](https://link.springer.com/book/10.1007/978-0-387-84858-7)
- [Computer Age Statistical Inference](https://www.cambridge.org/core/books/computer-age-statistical-inference/E32C1911ED937D75CE159BBD21684D37). <br> Bradley Efron and Trevor Hastie (2016)
- [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/). <br> Richard McElreath (2020)
- [Regression and Other Stories](https://www.cambridge.org/highereducation/books/regression-and-other-stories/DD20DD6C9057118581076E54E40C372C#overview). <br> Andrew Gelman, Jennifer Hill, and Aki Vehtari (2020)
- [Data Analysis Using Regression and Multilevel/Hierarchical Models](http://www.stat.columbia.edu/~gelman/arm/). <br> Andrew Gelman, Jennifer Hill (2006)
- [Introduction to Probability](https://www.routledge.com/Introduction-to-Probability-Second-Edition/Blitzstein-Hwang/p/book/9781138369917). <br> Joseph K. Blitzstein and Jessica Hwang (2019)
- [Bayesian Modeling and Computation in Python](https://www.routledge.com/Bayesian-Modeling-and-Computation-in-Python/Martin-Kumar-Lao/p/book/9780367894368). <br> Osvaldo A. Martin, Ravin Kumar, and Junpeng Lao (2022)
- [Bayes Rules!](https://www.routledge.com/Bayes-Rules-An-Introduction-to-Applied-Bayesian-Modeling/Johnson-Ott-Dogucu/p/book/9780367255398). <br> Alicia A. Johnson, Miles Q. Ott, and Mine Dogucu (2022)
- ...
